---
title: "Introduction Dplyr in R for Data Manipulation"
author: Andy Pickering
output: 
  html_document: 
    toc: no
---




# dplyr has 5 main verbs that each do 1 thing:

- select
- arrange
- filter
- mutate
- summarise


I'll use the *mtcars* dataset for examples.
```{r}
library(dplyr)
head(mtcars)
```

# select 

- 'select' selects columns from a data frame.
- Same as SQL select

```{r select}


# select specified columns
df <- select(mtcars,mpg,cyl)
head(df,5)


```

```{r}
# select everything *but* a column
df <- select(mtcars,-cyl)
head(df,5)

```


```{r}

# select columns starting with x...
df <- select(mtcars,starts_with('d'))
head(df,5)
```



# arrange
- Sorts rows by one or more specified variables
- Similar to SQL *order by*

```{r}

df <- arrange(mtcars,mpg)
head(df)

```


# filter
- filters rows based on given critera
- SQL equivalent would be *where ...*

```{r}

df <- filter(mtcars,mpg>30)
head(df)
```


# mutate
- Create new columns (or alter existing columns)

```{r}

df <- mutate(mtcars,mpkm = mpg*1.60934)
df <- select(df,mpg,mpkm)
head(df)
```

# summarise()
- More commonly used with group-by (see below)

```{r}

df <- summarise(mtcars,avg_mpg=mean(mpg))
df
```



# The pipe!

- I'll be using the *pipe* operator in these examples. The pipe takes the object on the left side and inputs it as the 1st argument to the function on the right. 
- This eliminates the need for intermediate objects and can make code easier to read.

- Let's say we want to select only the mpg,cyl, and disp columns, then filter by cyl, then order by mpg.

## without pipe
```{r, eval=FALSE}

# without pipe
df <- select(mtcars,mpg,cyl,disp)
df <- filter(df,cyl==8)
df <- arrange(df,mpg)
head(df)  

```


## with pipe
```{r}

df <- mtcars %>% select(mpg,cyl,disp) %>% filter(cyl==8) %>% arrange(mpg)
head(df)

```

You can also pipe your results directly into ggplot! 
```{r}
library(ggplot2)
mtcars %>% select(mpg,cyl,disp) %>%
          filter(cyl==8) %>%
          arrange(mpg) %>%
          ggplot(aes(disp,mpg))+
          geom_point()
```



# group by
- Group by 1 or more variables, and perform operations (ie average/sum) on each group


## Example: Compute the average mpg for each number of cyl

```{r}
mtcars %>% group_by(cyl) %>% summarise(avg=mean(mpg))
```


# Joins
```{r}

```


# Resources

<http://dplyr.tidyverse.org/>

<https://www.datacamp.com/courses/dplyr-data-manipulation-r-tutorial>

<https://www.datacamp.com/courses/joining-data-in-r-with-dplyr>


# SessionInfo
```{r}
sessionInfo()
```


